Automatic Matching of Bullet Lands

### 2016-04-16 revision/response.Rmd

12. In the conclusions section the authors mention an assumption of ``uniqueness". In general forensic science have moved beyond these questions- see the work of Saks and Kaye for details.

***probably refers to articles such as ``how good is good enough?", books ``modern scientific evidence", "judging science" .... I am not sure that those are really necessary or helpful. It might be better to re-phrase the sentence in the conclusion.***

13. A discussion on how the authors view the use of their matching algorithm would be useful to the forensic community. For example are they intending to use this to build a closed set identification system similar to IAFIS or are them intending this to work in a manner similar to a biometric verification system? etc.

***HH: in my view we ourselves are not building a closed set identification system. The problem with existing closed set systems is that it is unclear how the matching is exactly done and what statistical properties it has (number of parameters, distribution of them, sensitivities). Repeatability and reproducibility is a huge issue in this area. 'Matching' is also not just a single step - it consists, as we have laid out in the paper, of a series of steps, which all deserve attention. While the reviewer pointed out to us to use 'standard methods' - there are no actual standards in place. There might be a set of alternatives, and some evidence as to which methods might work better in some situations than others, but there is a lack of an in-depth discussion of the alternatives at each step as well as an overall discussion. In this manuscript we are detailing one route to divvy up the process and provide the code for others to reproduce our findings.  ***

@@

Reviewer C:

14. there is a lack of knowledge of the existing literature on statistical work performed on tool marks and firearms. For example, the work performed by Nicholas Petraco, Fabiano Riva, or Alan Zheng in the past 5 years is not mentioned. Instead, old research projects are mentioned, which mostly focus on manual identification. The authors need to discuss the novelty of their method (and how it potentially addresses the shortcomings of these recent methods)

15.  The authors would need to study (or at least discuss extensively) the ability of their method to capture the variability in bullet profiles over time (at interval of several dozen of bullets fired by the same weapon, when the weapon has been cleaned, or when the bullet was degraded by hitting a target). In addition, a dataset of 10 firearms (even consecutively manufactured) does not provide any information on the rarity of a particular bullet profile, and therefore does not address the second probability statement. Alternatively, the proposed metrics could be tested in a database searching situation, which has not been done.

***At the time of the submission of this manuscript, the 35 bullets from Hamby set 252 were the only ones accessible to the public. They are part of Dr Zheng's efforts at NIST to provide a ballistics reference database. In the future more results will be available, but until we are unable to provide a data-informed discussion of the very important issues raised by the reviewer.***

@@

***DONE***

24. p. 3, Fig. 1:  The numbers on the y-axis are superimposed.

25. p. 3, l. -10:  Be explicit that grooves and striations are synonyms.

26. p. 7, l. 1:  Make this "... can be described in the form ...".

***DONE***

27. p. 7, l. 9:  Remove the comma after "here" in "... function used here, ...".

***DONE***

28. p. 7, l. -6:  I'm confused.  The text says "bullets 2 are shown" but the caption refers to bullet 2- which I think should be a specific land on a single bullet.

***DONE***

29. p. 12, l. -11:  The use of a colon to transition to a new paragraph seems odd.  I recommend just rewording so is a single paragraph and does not use a colon at all.  It is almost as if you are trying to quote rather than state.

***DONE***

30. p. 13, l. -3:  You write "Given that striation patterns are typically much larger, ..." and I think it would be more clear to write "Given that striation widths are typically much larger, ...".

***DONE***

@@

***DONE***

46. Lines 156-158: The sentence ``The blue area around the signature are 95% bootstrapped confidence intervals based on 1,000 bootstrap samples." is confusing. The term ``signature" is defined two sentences further. Is the sentence referring to Figure 4b? If yes, it should probably come at the end of this paragraph.

47. Figure 5: In this figure, where is the limit between bullet 1-5 and bullet 2-1?

***Light grey shading in the background was added to help with the distinction between the two flattened bullet lands.***

@@

48. Line 246: Replace ``this representations" with ``this representation" or ``these representations"

***DONE***

49. Caption of Figure 8: ``At a lag of -17 the correlation peaks, indicating the largest amount of agreement between the signatures." is not a complete sentence.

***DONE***
***HH: would a removal of the comma make it a complete sentence?***

### 2016-04-16 revision/response.Rmd

4. Random forests, as used and described on pages 18-21 of the paper comes out looking like a winner, but other than that success---worth reporting *somewhere*---, there is no new information on that method. No new insights. And certainly no comparisons with the tens of other things that could have been tried.

Reviewer B:

5. The authors should discuss their work in the context of Nicholas Petraco’s work as well as other ongoing research endeavors in impression and pattern evidence. See https://www.ncjrs.gov/pdffiles1/nij/grants/239048.pdf The alignment method should have been compared with a standard method- See Patraco’s NIJ report for a couple of examples. As presented the novel component of the paper is the ``twicing" for grooves combined with a loess fit to the remove curves from bullet images. I am not sure this is the authors intention. It would be interesting to compare this method with some of the classical methods from functional data analysis used for curve registration.

***Additional papers considered (not complete, and not yet included in the paper): 
chumbley:2010 
uses 1d profiles and discusses several different angles of extraction (no alignment necessary, no grooves and no flattening)

chu:2010
describe extracting mean profile of an area of the image after untwisting the image

bachrach:2002
general description of an automated matching system based on 3D imaging. 
Uses cross-correlation function for alignment.
***

### 2016-04-16 revision/response.Rmd

5. The authors should discuss their work in the context of Nicholas Petraco’s work as well as other ongoing research endeavors in impression and pattern evidence. See https://www.ncjrs.gov/pdffiles1/nij/grants/239048.pdf The alignment method should have been compared with a standard method- See Patraco’s NIJ report for a couple of examples. As presented the novel component of the paper is the ``twicing" for grooves combined with a loess fit to the remove curves from bullet images. I am not sure this is the authors intention. It would be interesting to compare this method with some of the classical methods from functional data analysis used for curve registration.

***Additional papers considered (not complete, and not yet included in the paper): 
chumbley:2010 
uses 1d profiles and discusses several different angles of extraction (no alignment necessary, no grooves and no flattening)

chu:2010
describe extracting mean profile of an area of the image after untwisting the image

bachrach:2002
general description of an automated matching system based on 3D imaging. 
Uses cross-correlation function for alignment.

petraco:2012
I am not sure that this book is the same as the report for the NIJ grant. This is a huge effort, and we should certainly cite it, but there are not a lot of actual results included or comparisons done.

riva:2014
automatic matching system based on 3d images of cartridge cases. Matching involves breech faces, but no striation marks.
***

### 2016-04-17 revision/response.Rmd

4. Random forests, as used and described on pages 18-21 of the paper comes out looking like a winner, but other than that success---worth reporting *somewhere*---, there is no new information on that method. No new insights. And certainly no comparisons with the tens of other things that could have been tried.

Reviewer B:

5. The authors should discuss their work in the context of Nicholas Petraco’s work as well as other ongoing research endeavors in impression and pattern evidence. See https://www.ncjrs.gov/pdffiles1/nij/grants/239048.pdf The alignment method should have been compared with a standard method- See Patraco’s NIJ report for a couple of examples. As presented the novel component of the paper is the ``twicing" for grooves combined with a loess fit to the remove curves from bullet images. I am not sure this is the authors intention. It would be interesting to compare this method with some of the classical methods from functional data analysis used for curve registration.

***Using the cross correlation function as a method of alignment is actually pretty standard - the approriate references are now included.
***

***Additional papers considered (not complete, and not yet included in the paper, but I think all of them should be cited): 
chumbley:2010 
uses 1d profiles and discusses several different angles of extraction (no alignment necessary, no grooves and no flattening)
chubley:2010 is a much more detailed version over David Faden's thesis***

***chu:2010
describe extracting mean profile of an area of the image after untwisting the image ***

***bachrach:2002
general description of an automated matching system based on 3D imaging. 
Uses cross-correlation function for alignment.***

***petraco:2012
I am not sure that this book is the same as the report for the NIJ grant. This is a huge effort, and we should certainly cite it, but there are not a lot of actual results included or comparisons done.***

***riva:2014
automatic matching system based on 3d images of cartridge cases. Matching involves breech faces, but no striation marks.
***

### 2016-04-17 imaging-paper.Rnw
added citation

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are quite prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lof$bulletname <- scrubPath(lof$bullet)
qplot(y, resid, group=bulletname, colour=bulletname, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
    \caption{Aligned signatures.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lofY$bulletname <- scrubPath(lofY$bullet)
qplot(y+min(lof$y), resid, group=bulletname, colour=bulletname, data=lofY, 
      geom="line") + xlab("y") +
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

>>>>

\subsection{Horizontal alignment}
Signatures of each of the two lands, 1-5 and 2-1,  in Figure~\ref{fig:manualmatch-rgl} are shown in Figure~\ref{fig:cross100} extracted at a height of $x = 100\mu m$. Striation marks show up in these representations as peaks and valleys.  The individual characteristics are quite prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in Fig~\ref{fig:crosscutX}) emphasizes the strong similarities. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lof$bulletname <- scrubPath(lof$bullet)
qplot(y, resid, group=bulletname, colour=bulletname, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
    \caption{Aligned signatures.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
lofY$bulletname <- scrubPath(lofY$bullet)
qplot(y+min(lof$y), resid, group=bulletname, colour=bulletname, data=lofY, 
      geom="line") + xlab("y") +
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(1, 1.1), legend.justification=c("right", "top"),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 taken at  heights of $x = 100\mu m$. A horizontal shift of the values of bullet~1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
\hh{For this alignment we use the cross-correlation function to find a maximal amount of agreement between the signatures~\cite{bachrach:2002, thompson:2013}.}
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0~$\mu m$ and about 2500~$\mu m$, 1.5625~$\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).
